---
title             : "Police stop and depressive symptoms: Examining moderating role of race"
shorttitle        : "Police stop and depressive symptoms"

author: 
  - name          : "Mohammad Hashim"
    affiliation   : "1"
    corresponding : yes 
    address       : "Department of Psychology"
    email         : "hashi205@umn.edu"
    role: 
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  
affiliation:
  - id            : "1"
    institution   : "University of Minnesota"


authornote: |
  A sample project for the course of Graduate Seminar in Psychology, led by Dr. Moin Syed.

abstract: |
  Police contact represents a detrimental stressor that may be associated with elevated depressive symptoms, particularly among marginalized groups. Using a simulated dataset of 500 participants, we examined whether experiencing a police stop predicted PHQ-9 depressive symptom severity, and whether this association was moderated by race. Results from a linear model showed that individuals who had been stopped by police reported substantially higher depressive symptoms, B = 13.15, SE = 0.68, t(498) = 19.42, p < .001, accounting for approximately R^2 = .43 of the variance. A second model including race and the police stop × race interaction demonstrated improved model fit (R^2 = .47). The interaction term was significant, B = -6.38, SE = 1.62, t(496) = -3.93, p < .001, suggesting that the psychological impact of police stops was stronger for BIPOC participants than for White participants. These findings highlight the potential role of race in shaping vulnerability to the mental health consequences of police contact.
  
keywords          : "police stop; psychopathology; black; race; legal system expsoure"
wordcount         : "X"

bibliography      : "r-references.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---
Police interactions, especially involuntary or intrusive stops, are increasingly recognized as significant stressors that may undermine mental health. A growing body of research shows that being stopped by the police can evoke fear, threat, and feelings of injustice, all of which may contribute to elevated depressive symptoms. However, the psychological impact of police contact is not experienced uniformly across communities. Race remains a central factor shaping how individuals perceive, interpret, and internalize police encounters. For many racial and ethnic minority groups, especially Black and Latino communities, police stops occur within a broader historical and social context marked by discrimination and disproportionate surveillance.
The present study examines the association between police stops and depressive symptoms and investigates whether this relationship differs by race. Understanding racial variation in the mental health consequences of police contact is essential for clarifying risk pathways and identifying populations most adversely affected. This work contributes to ongoing discussions on policing, public health, and racial inequality by evaluating whether race moderates the psychological burden of police stops.


```{r setup, include = false}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```



# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.


# Results

# Discussion


\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
